HealpixNet: Efficient spherical Convolutional Neural Network with HEALPix sampling for cosmological applications

### 2018-08-10 scnn.tex
small edit and merge

\section*{TODO list}
\begin{itemize}[noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt]
    \item (TK) check text figure 1
	\item (TK) Related work in cosmo. The only cosmological applications of spherical CNNs in the related work are about projection on the tangent plane \citep{schmelze2017cosmologicalmodel, fluri2018deep, gupta2018nongaussianinformation}. Aren't there applications in cosmology for the other (c.f. related work section) approaches? Most of them are illustrated in omnidirectional imagery.
	\item (MD) distinction between CNN and FCN
    \item (first version done!)(NP+MD) paragraph on our chosen architecture in the Experiments sections
    \item (MD) Are you happy with the vertex interpretation in section 2.5 Efficient convolutions?
    \item (TK) check if papers which appear as arxiv were not published in a conference or journal
    \item (TK) check part sphere notebook and whole sphere
    \item (TK) read everything
    \item (NP) Put Raphael in the dataset author. 
    \item (done!) (TK) add small description of maps\_downsampled\_64.npz in the whole\_sphere notebook. \nati{We need to talk about that.}
    \item (TK) color text in 2.8
    \item (NP) D.18 text size bigger
    \item (NP) Remove caption B.13
	\item Beware to not confuse reader with synonyms, e.g. pixelization and sampling.
    \item Check names: HealpixNet, HealpixCNN, HealpixFCN
    \item (NP) Check the wording. Be sure that the convolution on SO(3) is clear: we want to say (a) is the one associated with the SO(3) rotation group (b) can be performed  in the spectral domain by a multiplication (c) is equivariant to rotation (d) is described well in Section 2 of \citep{kondor2018clebsch}.
    \item Change equation after (2) and check all kroneker
	\item Notation: write element $i$ of vector $y$ as $y_i$, $y(i)$, or $y[i]$? Be consistent. Same for matrices.
	\item Try to make the paper shorter?
	\item (TK) Font: do we have to use times (set in the documentclass)? I (MD) find the rendering of math a bit ugly.
    \item (TK) fix the bibliography
    \item Check links in footnotes.
\end{itemize}

\begin{figure*}
	\centering
    % \includegraphics[width=\linewidth]{architecture}
	\includegraphics[width=\linewidth]{figure_architecture_v3.pdf}
	\caption{Overall network architecture, showing here two convolutional layers acting as feature extractors followed by a fully connected layer with softmax acting as the classifier.
    %\nati{add the stat layer?}
    %\TK{added stats layer to figure text, hope that's OK}
    A convolutional layer is based on five operations: convolution, non-linearity, batch normalization, down-sampling, and pooling. While most operations are agnostic to the data domain, the convolution and the down-sampling have to be adapted. In this paper, we propose first to model the sphere with a graph and to perform the convolution on the graph. Graphs are versatile data structures which can model any sampling, even irregular or partial. Second, we propose to exploit a hierarchical pixelization of the sphere for the pooling operation. It allows the network to analyze the data at multiple scales while preserving the spatial localization of features. This figure shows a network which operates on the whole sphere. The process is however the same when working with parts of sphere, except that the graph is only built for the part of interest.}
	\label{fig:architecture}
\end{figure*}

@@

So far these algorithms have mostly been demonstrated on Euclidean domains, such as flat images.
Perhaps the most straightforward approach to the analysis of spherical data is to divide it into small chunks and project those on flat 2D surfaces. \todo{Tune according to the story in the related work.}
While this approach has been demonstrated to work \citep{fluri2018deep,gupta2018nongaussianinformation,schmelze2017cosmologicalmodel,gillet2018deeplearning}, it may be more natural for some applications and accurate to work directly on the sphere.
The main challenge in designing a CNN on the sphere is to define a convolution operation that is suitable for this irregularly sampled domain.
To be able to train the network efficiently, the convolution has to be performed fast.
Another important feature is for a CNN to be able to work on part-spheres, as for many cosmological observations cover only a part of the sky.
For ground-based observations this can be due to limited visibility of the sky from a particular telescope location, and for space-based instruments due to masking of the galactic plane area.

@@

\citet{cohen2018sphericalcnn} and \citet{esteves2017sphericalcnn} addressed the problem of rotation equivariance by defining the convolution on the 3D rotation group SO(3), with applications to atomization energy regression and 3D model classification, alignment and retrieval.
The resulting convolution is performed by (i) a spherical harmonic transform (SHT), i.e., a projection on the spherical harmonics, (ii) a  multiplication in the spectral domain, and (iii) an inverse SHT. Hence the computational cost of a convolution is dominated by the two SHTs.
A naive implementation of the SHT costs $\bO(N_{pix}^2)$ operations, where $N_{pix}$ is the number of pixels. Note the similarity with the naive graph convolution defined in \eqnref{graph_convolution_fourier}.
Accelerated schemes exist for some sampling sets \citep[see][for examples]{mohlenkamp1999fast, rokhlin2006fast, reinecke2013libsharp}.
The main advantage of this approach is that it provides a mathematically well defined rotation-equivariant network. Nevertheless, even with accelerated SHTs, the convolutions remain expensive, limiting the practical use of this approach.
For example with the HEALPix sampling, which was designed to have a fast SHT by being iso-latitude, the computational cost of the SHT is $\bO(N_{pix}^{3/2}) = \bO(N_{side}^3) = \bO(\ell_{max}^3)$, where $\ell_{max}$ is the largest angular frequency\footnote{All pixels are placed on $N_{ring} = 4N_{side}-1 = \bO\left(\sqrt{N_{pix}}\right)$ rings of constant latitude. Each ring has $\bO\left(\sqrt{N_{pix}}\right)$ pixels. Thanks to this iso-latitude property, the SHT is computed using recurrence relations for Legendre polynomials on co-latitude and fast Fourier transforms (FFTs) on longitude. The computational cost is thus $\sqrt{N_{pix}}$ FFTs for a total cost of $\bO\left( N_{pix} \log \sqrt{N_{pix}} \right)$, plus $\sqrt{N_{pix}}$ matrix multiplications of size $\sqrt{N_{pix}}$ for a total cost of $\bO\left(N_{pix}^{3/2}\right)$.}
\citep{gorski2005healpix, reinecke2013libsharp}.
In comparison, filtering with our HealpixNet method scales as $\bO(N_{pix})$ (see \secref{efficient_convolution}).
\citet{kondor2018clebsch} proposed to use the Clebsh-Gordan transform to perform a non-linear transformation in the spectral domain. As such, a SHT only has to be performed once to transform the data in the spectral domain. No need to transform back and forth between every layer. While a clever trick, it is unclear if it results in any speed gain in practice. \todo{computation complexity? no speed comparison in paper?}
\todo{Polynomial formulation enforces locality, which otherwise has to be enforced by smoothness in the spectral domain.}
\todo{Pooling in the spectral domain? \citet{esteves2017sphericalcnn} do something.}
% While a comparison is theoretically possible, the experiments of~\cite{cohen2018sphericalcnn} are done using the geodesic grid sampling set and do not provide any code for the HEALPix one. Hence, a practical comparison is not possible.

@@

\nati{Michael: I am highly uncertain of what should be kept in the following paragraph. Maybe the full paragraph should be left out!}
While testing the FCN and CNN architectures, we made the following empirical observations.
First, Chebyshev polynomial seems more stable than monomials ($\theta_k$ in \eqnref{graph_convolution_polynomial}), in the sense that the training loss had a more monotone decrease.
Nevertheless once the network is trained, we could not observe a significant accuracy difference.
Second, using a statistical layer (FCN) to get equivariant features slightly improved the accuracy in our problem.
We tried $4$ different layers `mean', `variance', `mean + variance' and `histogram'.
The `mean', `histogram' and `mean + variance' were providing the same level of accuracy, with pure `variance' being slightly worse. Hence we chose to simply use the `mean' statistic.
Third, adding fully connected layers after the statistical layer does not improve the performance, but lead to models requiring more training steps and regularization.
Fourth, using $\ell_2$ regularization does not help either with the performance or the training of the models.
Fifth, we recommend initializing the Chebyshev coefficients with a centered Gaussian distribution with standard deviation $\sqrt{2/(F * (K + 0.5))}$, where $K$ is the Chebyshev order and $F$ the number of input channels. This standard deviation has the property to keep the energy of the signal more or less at the same level between the different layers.
% Nati to Michael: The problem with the previous claim is that it depends on the non-linarity. Furthermore, does it make sense to make such a claim if we use batch norm?
Eventually, we observe that rescaling the Laplacian eigenvalues between $[-a,a]$, where $1<a\leq1$ significantly helps to stabilize the optimization. For our experiment we use $a=0.75$.

>>>>

\section*{TODO list}
\begin{itemize}[noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt]
    \item (TK) check text figure 1
	\item (TK) Related work in cosmo. The only cosmological applications of spherical CNNs in the related work are about projection on the tangent plane \citep{schmelze2017cosmologicalmodel, fluri2018deep, gupta2018nongaussianinformation}. Aren't there applications in cosmology for the other (c.f. related work section) approaches? Most of them are illustrated in omnidirectional imagery.
	\item (MD) distinction between CNN and FCN
    \item (first version done!)(NP+MD) paragraph on our chosen architecture in the Experiments sections
    \item (MD) Are you happy with the vertex interpretation in section 2.5 Efficient convolutions?
    \item (TK) check if papers which appear as arxiv were not published in a conference or journal
    \item (TK) check part sphere notebook and whole sphere
    \item (TK) read everything
    \item (NP) Put Raphael in the dataset author. 
    \item (done!) (TK) add small description of maps\_downsampled\_64.npz in the whole\_sphere notebook. \nati{We need to talk about that.}
    \item (TK) color text in 2.8
    \item (NP) D.18 text size bigger
    \item (NP) Remove caption B.13
	\item Beware to not confuse reader with synonyms, e.g. pixelization and sampling.
    \item Check names: we use HealpixNet, HealpixNet variant CNN, HealpixNet variant FCN
    \item (NP) Check the wording. Be sure that the convolution on SO(3) is clear: we want to say (a) is the one associated with the SO(3) rotation group (b) can be performed  in the spectral domain by a multiplication (c) is equivariant to rotation (d) is described well in Section 2 of \citep{kondor2018clebsch}.
    \item Change equation after (2) and check all kroneker
	\item Notation: write element $i$ of vector $y$ as $y_i$, $y(i)$, or $y[i]$? Be consistent. Same for matrices.
	\item Try to make the paper shorter?
	\item (TK) Font: do we have to use times (set in the documentclass)? I (MD) find the rendering of math a bit ugly.
    \item (TK) fix the bibliography
    \item Check links in footnotes.

\end{itemize}

\begin{figure*}
	\centering
    % \includegraphics[width=\linewidth]{architecture}
	\includegraphics[width=\linewidth]{figure_architecture_v3.pdf}
	\caption{Overall network architecture, showing here two convolutional layers acting as feature extractors followed by a fully connected layer with softmax acting as the classifier.
    %\nati{add the stat layer?}
    %\TK{added stats layer to figure text, hope that's OK}
    A convolutional layer is based on five operations: convolution, non-linearity, batch normalization, down-sampling, and pooling. While most operations are agnostic to the data domain, the convolution and the down-sampling have to be adapted. In this paper, we propose first to model the sphere with a graph and to perform the convolution on the graph. Graphs are versatile data structures which can model any sampling, even irregular or partial. Second, we propose to exploit a hierarchical pixelization of the sphere for the pooling operation. It allows the network to analyze the data at multiple scales while preserving the spatial localization of features. This figure shows a network which operates on the whole sphere. The process is, however, the same when working with parts of sphere, except that the graph is only built for the part of interest.}
	\label{fig:architecture}
\end{figure*}

@@

So far these algorithms have mostly been demonstrated on Euclidean domains, such as flat images.
Perhaps the most straightforward approach to the analysis of spherical data is to divide it into small chunks and project those on flat 2D surfaces. \todo{Tune according to the story in the related work.}
While this approach has been demonstrated to work \citep{fluri2018deep,gupta2018nongaussianinformation,schmelze2017cosmologicalmodel,gillet2018deeplearning}, it may be more natural for some applications and accurate to work directly on the sphere.
The main challenge in designing a CNN on the sphere is to define a convolution operation that is suitable for this irregularly sampled domain.
To be able to train the network efficiently, the convolution has to be performed fast.
Another important feature is for a CNN to be able to work on subpart of the sphere, as for many cosmological observations cover only a part of the sky.
For ground-based observations this can be due to limited visibility of the sky from a particular telescope location, and for space-based instruments due to masking of the galactic plane area.

@@

\citet{cohen2018sphericalcnn} and \citet{esteves2017sphericalcnn} addressed the problem of rotation equivariance by defining the convolution on the 3D rotation group SO(3), with applications to atomization energy regression and 3D model classification, alignment and retrieval.
The resulting convolution is performed by (i) a spherical harmonic transform (SHT), i.e., a projection on the spherical harmonics, (ii) a  multiplication in the spectral domain, and (iii) an inverse SHT. Hence the computational cost of a convolution is dominated by the two SHTs.
A naive implementation of the SHT costs $\bO(N_{pix}^2)$ operations, where $N_{pix}$ is the number of pixels. Note the similarity with the naive graph convolution defined in \eqnref{graph_convolution_fourier}.
Accelerated schemes exist for some sampling sets \citep[see][for examples]{mohlenkamp1999fast, rokhlin2006fast, reinecke2013libsharp}.
The main advantage of this approach is that it provides a mathematically well defined rotation-equivariant network. Nevertheless, even with accelerated SHTs, the convolutions remain expensive, limiting the practical use of this approach.
For example with the HEALPix sampling, which was designed to have a fast SHT by being iso-latitude, the computational cost of the SHT is $\bO(N_{pix}^{3/2}) = \bO(N_{side}^3) = \bO(\ell_{max}^3)$, where $\ell_{max}$ is the largest angular frequency\footnote{All pixels are placed on $N_{ring} = 4N_{side}-1 = \bO\left(\sqrt{N_{pix}}\right)$ rings of constant latitude. Each ring has $\bO\left(\sqrt{N_{pix}}\right)$ pixels. Thanks to this iso-latitude property, the SHT is computed using recurrence relations for Legendre polynomials on co-latitude and fast Fourier transforms (FFTs) on longitude. The computational cost is thus $\sqrt{N_{pix}}$ FFTs for a total cost of $\bO\left( N_{pix} \log \sqrt{N_{pix}} \right)$, plus $\sqrt{N_{pix}}$ matrix multiplications of size $\sqrt{N_{pix}}$ for a total cost of $\bO\left(N_{pix}^{3/2}\right)$.}
\citep{gorski2005healpix, reinecke2013libsharp}.
In comparison, filtering with our HealpixNet method scales as $\bO(N_{pix})$ (see \secref{efficient_convolution}).
\citet{kondor2018clebsch} proposed to use the Clebsh-Gordan transform to perform a non-linear transformation in the spectral domain. As such, a SHT only has to be performed once at the input of the network and the back and forth SHT between every layer are no longer required. While this clever trick make the convolution complexity $\bO(N_{pix})$, the non-linearity\nati{ scales as the squared of the number of channels and as }. As a result, it is yet unclear which speed gain is obtained in practice. \todo{computation complexity? no speed comparison in paper?}
\todo{Polynomial formulation enforces locality, which otherwise has to be enforced by smoothness in the spectral domain.}
\todo{Pooling in the spectral domain? \citet{esteves2017sphericalcnn} do something.}
% While a comparison is theoretically possible, the experiments of~\cite{cohen2018sphericalcnn} are done using the geodesic grid sampling set and do not provide any code for the HEALPix one. Hence, a practical comparison is not possible.

@@

\nati{Michael: I am highly uncertain of what should be kept in the following paragraph. Maybe the full paragraph should be left out!}
While testing the FCN and CNN architectures, we made the following empirical observations.
First, Chebyshev polynomial seems more stable than monomials ($\theta_k$ in \eqnref{graph_convolution_monomial}), in the sense that the training loss had a more monotone decrease.
Nevertheless once the network is trained, we could not observe a significant accuracy difference.
Second, using a statistical layer (FCN) to get equivariant features slightly improved the accuracy in our problem.
We tried $4$ different layers `mean', `variance', `mean + variance' and `histogram'.
The `mean', `histogram' and `mean + variance' were providing the same level of accuracy, with pure `variance' being slightly worse. Hence we chose to simply use the `mean' statistic.
Third, adding fully connected layers after the statistical layer does not improve the performance, but lead to models requiring more training steps and regularization.
Fourth, using $\ell_2$ regularization does not help either with the performance or the training of the models.
Fifth, we recommend initializing the Chebyshev coefficients with a centered Gaussian distribution with standard deviation $\sqrt{2/(F * (K + 0.5))}$, where $K$ is the Chebyshev order and $F$ the number of input channels. This standard deviation has the property to keep the energy of the signal more or less at the same level between the different layers.
% Nati to Michael: The problem with the previous claim is that it depends on the non-linarity. Furthermore, does it make sense to make such a claim if we use batch norm?
Eventually, we observe that rescaling the Laplacian eigenvalues between $[-a,a]$, where $1<a\leq1$ significantly helps to stabilize the optimization. For our experiment we use $a=0.75$.
